The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-15 WavesData-2019/LatexNotes/notes.tex
NM

% Basic physical parameters and scales
\newcommand{\depth}{h}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\freqp}{f_p}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamfac}{N}
% More parameters
\newcommand{\etastd}{\eta_{\text std}}
\newcommand{\amp}{a}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
\newcommand{\drat}{D}
\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Hp}{H^{+}}
\newcommand{\Hm}{H^{-}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gp}{\Gibbs^{+}}
\newcommand{\Gm}{\Gibbs^{-}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

% Nondimensionalization
\subsection{Dimensionless variables}
\label{nondim}
We introduce the following characteristic scales
\begin{align}
\label{ampdefn}
&\amp = \mean{\eta^2}^{1/2} 
&&\mbox{\em characteristic wave amplitude} \\
\label{lamdefn}
&\lam = \sqrt{g \depth} / \freqp
&&\mbox{\em characteristic wave length}
\end{align}
where $\mean{\cdot}$ indicates the mean of a quantity. Note that the characteristic wavelength $\lam = \lam(x)$ takes different values upstream and downstream of the ADC.
We introduce the following dimensionless variables
\begin{align}
&u = \eta/\amp
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{2 \pi \xi}{\lamfac \lam} = \frac{2 \pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time} \\
\end{align}
We will consider the physical domain $\xi \in [0, \lamfac \lam]$, which is an integer multiple of the characteristic wavelength where $\lamfac$ will be selected later (probably about 5). Above, we have chosen the scale to normalize position as $\lamfac \lam/(2\pi)$ so that the dimensionless domain can be taken (after a shift) to be $\tilde{x} \in [-\pi,\pi]$. Accordingly, we have also chosen the time scale as $\lamfac/\freqp$, so that a wave traveling at the local wavespeed $c = \lam \freqp = \sqrt{g \depth}$ crosses the physical domain $x \in [0, \lamfac \lam]$ in one unit of dimensionless time. Said differently, the dimensionless wave speed is $(2 \pi)^{-1}$ so that a wave crosses the entire domain $[-\pi,\pi]$ in unit time.

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
u_t + {3 \pi} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat(x) = {\depth(x)}/{\dup} \, .
\end{align}
Here, $\drat = \drat(x)$ is the local dimensionless depth. See Table \ref{paramtable} for a summary of important parameters.

@@

\subsection{Hamiltonian analysis}

The domain of consideration is $x \in [-\pi,\pi]$. First, the spatial average of any quantity is given by
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Appealing to near Ergodicity, we will typically assume equivalence between spatial, temporal, and ensemble averages. Note that from the definition of the characteristic amplitude in \eqref{ampdefn}, we have $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
E[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}

\vsp{10}

Then the Hamiltonian is given by
\begin{align}
\label{Hamiltonian}
& \Hm = \ampp H_3 - \mu H_2 \\
& \Hp = \ampp D^{-3/2} H_3 - \mu D^{1/2} H_2
\end{align}
where
\begin{align}
&H_3 = \frac{1}{6} \intt u^3 \dx	\, , \\
&H_2 = \frac{1}{2} \intt u_x^2 \dx	\, .
\end{align}

Note: after giving it some deliberation, I believe we should use $\lamfac=1$. My reason is that the peak in the wave spectrum occurs at a frequency of $f_p$ or wavelength of $\lambda_p$. In the tKdV Gibbs measure, the spectrum decays monotonically, so that the peak is at the lowest resolved frequency (or largest resolved wavelength). Thus, we want $\lambda_p$ to correspond to the largest resolved wavelength, i.e.~the length of the periodic domain in the tKdV framework.

%Remark: We note that another possibility is to match the quantity $D^{1/4} \eta$ at the abrupt depth change as would be consistent with the analysis of Johnson. If we were to do this, we would obtain only slightly different powers of $D$ in the coefficients: $D^{-7/4}$ and $D^{1/2}$ respectively.

NOTE: We need to include the bandwidth, 2 Hz, in the theory somehow. For example, the peak forcing frequency and bandwidth together can be used to estimate the decay in the spectrum.

>>>>

% Basic physical parameters and scales
\newcommand{\depth}{h}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\freqp}{f_p}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamfac}{N}
% More parameters
\newcommand{\etastd}{\eta_{\text std}}
\newcommand{\amp}{a}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
\newcommand{\drat}{D}
\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Ham}{H}
\newcommand{\Hp}{\Ham^{+}}
\newcommand{\Hm}{\Ham^{-}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gp}{\Gibbs^{+}}
\newcommand{\Gm}{\Gibbs^{-}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

% Nondimensionalization
\subsection{Dimensionless variables}
\label{nondim}

We introduce the following characteristic scales
\begin{align}
\label{ampdefn}
&\amp = \mean{\eta^2}^{1/2} 
&&\mbox{\em characteristic wave amplitude} \\
\label{lamdefn}
&\lam = \sqrt{g \depth} / \freqp
&&\mbox{\em characteristic wave length}
\end{align}
where $\mean{\cdot}$ indicates the mean of a quantity. Note that the characteristic wavelength $\lam = \lam(x)$ takes different values upstream and downstream of the ADC. We remark that experimental measurements indicate that the characteristic amplitude $\amp = \mean{\eta^2}^{1/2}$ is nearly the same on both values of the ADC. Hence, we will not distinguish between upstream and downstream values of $\amp$.
We introduce the following dimensionless variables
\begin{align}
&u = \eta/\amp
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{2 \pi \xi}{\lamfac \lam} = \frac{2 \pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time} \\
\end{align}
We will consider the physical domain $\xi \in [0, \lamfac \lam]$, which is an integer multiple of the characteristic wavelength where $\lamfac$ will be selected later (probably about 5). Above, we have chosen the scale to normalize position as $\lamfac \lam/(2\pi)$ so that the dimensionless domain can be taken (after a shift) to be $\tilde{x} \in [-\pi,\pi]$. Accordingly, we have also chosen the time scale as $\lamfac/\freqp$, so that a wave traveling at the local wavespeed $c = \lam \freqp = \sqrt{g \depth}$ crosses the physical domain $x \in [0, \lamfac \lam]$ in one unit of dimensionless time. Said differently, the dimensionless wave speed is $(2 \pi)^{-1}$ so that a wave crosses the domain $[-\pi,\pi]$ in unit time.

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV}
u_t + {3 \pi} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat(x) = {\depth(x)}/{\dup} \, .
\end{align}
Here, $\drat = \drat(x)$ is the local dimensionless depth. See Table \ref{paramtable} for a summary of important parameters.

@@

\subsection{Hamiltonian structure}

We now consider the Hamiltonian structure of the dimensionless variable-depth KdV equation \eqref{dimlessKdV}, over dimensionless domain $x \in [-\pi,\pi]$. First, the spatial average of any quantity is given by
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Appealing to near Ergodicity, we will typically assume equivalence between spatial, temporal, and ensemble averages.

We measure the state variable $u$ as the displacement of the surface from equilibrium, so that $\mean{u} = 0$. Consequently, the momentum of $u$ vanishes
\begin{equation}
M[u] \defeq \int_{-\pi}^{\pi} u \dx = 0
\end{equation}
Next, the definition of the characteristic amplitude in \eqref{ampdefn} implies that $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
E[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}

We introduce the cubic and quadratic Hamiltonian components as
\begin{align}
H_3 = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
H_2 = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}

Then the Hamiltonian of \eqref{dimlessKdV} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}

\begin{comment}
Old Note (around May 2019): After giving it some deliberation, I believe we should use $\lamfac=1$. My reason is that the peak in the wave spectrum occurs at a frequency of $\freqp$ or wavelength of $\lam$. In the tKdV Gibbs measure, the spectrum decays monotonically, so that the peak is at the lowest resolved frequency (or largest resolved wavelength). Thus, we want $\lam$ to correspond to the largest resolved wavelength, i.e.~the length of the periodic domain in the tKdV framework.

New Thought (July 2019): Actually, it is very possible that the bandwidth (2 Hz in the experiments) sets the value of $\lamfac$ that is most appropriate, since that bandwidth sets the decay rate of the spectrum. My thinking is that our experimental forcing does not really follow the upstream Gibbs measure all that well (because we did not have that in mind), but perhaps it best approximates the tail decay of some Gibbs measure. Perhaps, the lowest frequencies (i.e. those much slower than 2 Hz that are present in the theory but almost absent in the experiments) do not affect statistics that much.
\end{comment}
